Recent demography drives changes in linked selection across the maize genome

### 2015-11-06 manuscript.tex
Final... (yeah right!)

\begin{figure*}[!tb]
%\vspace*{.05in}
\centering
\includegraphics[width=.4\textwidth]{FigsAndFiles/DomesticationModel/domesticationModel.pdf}
\caption{Parameter estimates for a basic bottleneck model of maize domestication. See methods for details. \label{fig:bottleneck} }
\end{figure*}

@@

In the case of purifying or background selection, diversity can be reduced in functional regions of the genome via removal of deleterious mutations \cite{charlesworth1993}.
We investigated purifying selection in maize and teosinte by evaluating the reduction of diversity around genes.
Pairwise diversity is strongly reduced within genes for both maize and teosinte (Figure \ref{fig:purify}A) but recovers quickly at sites outside of genes, consistent with the low levels of linkage disequilibrium generally observed in these subspecies \cite{tenaillon2002,chia2012}. 
The reduction in relative diversity is more pronounced in teosinte, reaching lower levels in genes and occurring over a wider region.

Our comparison of synonymous and missense substitutions has low power to detect the effects of selection acting on multiple mutations or standing genetic variation, because in such cases diversity is not necessarily reduced \cite{innan2004,messer2013}. 
Such ``soft sweeps'' are still expected to occur more frequently in functional regions of the genome and could provide an alternative explanation for the observed reduction of diversity in genes. 
To test this possibility, we performed a genome-wide scan for selection using the H12 statistic, a method expected to be sensitive to both hard and soft sweeps \cite{garud2015}. 
Qualitative differences between maize and teosinte remained unchanged even after removing genes in the top 20\% of H12 (Figure \ref{sFig:H12}A).
%\mbh{This makes me wonder about the top 20\%. Is the pattern of diversity reduction different in this fraction of the genome?  If so, that seems like something that should be mentioned.}
%TB Says: Sites in the top 20% of H12 are unlikely to be under background selection, so maize vs teo background selection pattern at these sites seems strange to investigate.
We interpret these combined results as suggesting that purifying selection has left a more pronounced signature in the teosinte genome due to the increased efficacy of selection resulting from differences in effective population size.

\begin{figure*}[!tb]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative diversity versus distance to nearest gene in maize and teosinte. 
Shown are \textbf{A} pairwise nucleotide diversity and \textbf{B} singleton diversity.  
Relative diversity is calculated compared to the mean diversity in windows $\geq 0.01 cM$ or $\geq 0.02 cM$ from the nearest
gene for pairwise diversity and singletons, respectively. 
  Lines depict cubic smoothing splines with smoothing parameters chosen via generalized cross validation and shading depicts bootstrap-based 95\% confidence intervals.
  Inset plots depict a smaller range on the x-axis. \label{fig:purify}
  }
\end{figure*}

\subsection{Population expansion leads to stronger purifying selection in modern maize} %JRI
Motivated by the rapid post-domestication expansion of maize evident in our demographic analysis, we reasoned that low-frequency --- and thus younger --- polymorphisms might show patterns distinct from pairwise diversity. 
Singleton diversity around missense and synonymous substitutions (Figure \ref{sFig:singleton}) appears nearly identical to results from pairwise diversity (Figure \ref{fig:hardSweeps}), providing little support for a substantial increase in the number or strength of hard sweeps occurring in maize. 
In contrast, we observe a significant shift in the effects of purifying selection: singleton polymorphisms are more strongly reduced in and near genes in maize than in teosinte (Figure \ref{fig:purify}B), even after downsampling our maize data to account for differences in sample size. This result is the opposite of the pattern observed for $\pi$, where teosinte demonstrated a stronger reduction of diversity in and around genes than did maize. As was the case when we analyzed $\pi$, this relationship remained after we removed the 20\% of genes with the highest H12 values (Figure \ref{sFig:H12}).
Finally, while direct comparison of pairwise and singleton diversity within taxa are consistent with at least some non-equilibrium dynamics in teosinte, as indicated by our coalescent-based demographic analysis, these too reveal much stronger differences in maize (Figure \ref{sFig:singletonPi}).

@@

\begin{figure*}[!tb]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvM_Folded2_Significance_Aug}
\hspace{0.05\textwidth} \includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvT_Folded2_Significance_Aug}
\caption{Pairwise diversity surrounding synonymous and missense  substitutions in {\bf A} maize and {\bf B} teosinte. Axes show absolute diversity values (right) and values relative to mean nucleotide diversity in windows $\geq 0.01 cM$ from a substitution (left).  Lines depict a loess curve (span of 0.01) and shading represents bootstrap-based 95\% confidence intervals. Inset plots depict a larger range on the x-axis. \label{fig:hardSweeps}}
\end{figure*}

@@

@@

@@

\subsection{Demography of domestication} %JRI
Although a number of authors have investigated the demography of maize domestication \cite{eyre1998, tenaillon2004, wright2005}, these  efforts relied on data only from genic regions of the genome and made a number of limiting assumptions about the demographic model.  We show that diversity within genes has been strongly reduced by the effects of linked selection, such that even synonymous polymorphisms in genes are not representative of diversity at unconstrained sites. This implies that genic polymorphism data are unable to tell the complete or accurate demographic history of maize, but the rapid recovery of diversity outside of genes demonstrates that sites far from genes can be reasonably used for demographic inference. Furthermore, by utilizing the full joint SFS, we are able to estimate population growth, gene flow, and the strength of the domestication bottleneck without requiring assumptions about its duration.

One surprising result from our model is the estimated timing of domestication at $\approx 15,000$ years before present.
While this appears to conflict with archaeological estimates \cite{piperno2009}, we emphasize that this estimate reflects the fact that the genetic split between populations likely preceded anatomical changes that can be identified in the archaeological record. 
We also note that our result may also be inflated due to population structure, as our geographically diverse sample of teosinte may include populations diverged from those that gave rise to maize.

\subsection{Demography influences the efficiency of purifying selection}
One of our more striking findings is that the impact of purifying selection on maize and teosinte variability has varied through time. Although we observe a more pronounced decrease in $\pi$ around genes in teosinte than maize (Figure \ref{fig:purify}A), we see a greater decrease in singleton diversity in maize than teosinte (Figure \ref{fig:purify}B). Purifying selection is more efficient in larger populations  \cite{kimura1984}, and our observations using $\pi$  are thus consistent with our demographic model and previous work \cite{eyre1998, tenaillon2004, wright2005, ross2009} which shows a smaller long-term effective population size in maize. Similarly, our observations based on singleton diversity, which thereby reflect more recent patterns of selection, again are in agreement with maize and teosinte demography; from the viewpoint of a recent mutation, the rapid expansion of maize post-domestication means that today maize has a vastly larger effective population size than does teosinte.

A consequence of the previously-inefficient purifying selection in maize is that it likely harbors more weakly deleterious alleles segregating at intermediate frequency than does teosinte. Therefore, high genetic load could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. Additionally, the large $N_e$ of modern maize compared to modern teosinte implies that for new mutations, selection will operate much more efficiently in maize. This observation is not restricted to maize. Bottlenecks have been documented or predicted across most domesticated plant and animal species \cite{doebley2006, hyten2006, zhu2007, bovine2009, larson2013 }, but rapid expansion post-bottleneck, which we demonstrate generates ubiquitous consequences relating to genetic diversity, is not well studied. In human, model, and other species, these results demonstrate that observed patterns of selection depend heavily on recent and/or ancient patterns of demography.

%\jri{these results have implications for lots of studies. recent demographic history often ignored (examples). in crops, bneck is considered and loss of diversity, but expansion also likely ubiquitous and consequences not well studied.  in humans XYZ, and we see some of that already here. this argues that a full understanding of diversity in both wild and cultivated plants will require recent demography. mention C-D ignored recent demography -- here's where rec/pi windows could help.}

%\jri{I would like to see a supplemental figure or maybe just number here on the partial correlation in both teosinte and maize of diversity with recombination after accounting for gene density.  do this in 500kb windows and then in 10kb windows.  we should do the correlation with both pi and singletons.  This let's us compare more explictly with studies like C-D who evaluate overall patterns of linked selection. will be especially cool if we don't see a big difference in big windows but we do in 10kb windows -- siggests C-D looked at wrong scale  }

>>>>

\begin{figure*}[!tb]
%\vspace*{.05in}
\centering
\includegraphics[width=.5\textwidth]{FigsAndFiles/DomesticationModel/domesticationModel.pdf}
\caption{Parameter estimates for a basic bottleneck model of maize domestication. See methods for details. \label{fig:bottleneck} }
\end{figure*}

@@

@@

\begin{figure*}[tb]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvM_Folded2_Significance_Aug}
\hspace{0.05\textwidth} \includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvT_Folded2_Significance_Aug}
\caption{Pairwise diversity surrounding synonymous and missense  substitutions in {\bf A} maize and {\bf B} teosinte. Axes show absolute diversity values (right) and values relative to mean nucleotide diversity in windows $\geq 0.01 cM$ from a substitution (left).  Lines depict a loess curve (span of 0.01) and shading represents bootstrap-based 95\% confidence intervals. Inset plots depict a larger range on the x-axis. \label{fig:hardSweeps}}
\end{figure*}

@@

In the case of purifying or background selection, diversity is reduced in functional regions of the genome via removal of deleterious mutations \cite{charlesworth1993}.
We investigated purifying selection in maize and teosinte by evaluating the reduction of diversity around genes.
Pairwise diversity is strongly reduced within genes for both maize and teosinte (Figure \ref{fig:purify}A) but recovers quickly at sites outside of genes, consistent with the low levels of linkage disequilibrium generally observed in these subspecies \cite{tenaillon2002,chia2012}. 
The reduction in relative diversity is more pronounced in teosinte, reaching lower levels in genes and occurring over a wider region.

Our previous comparison of synonymous and missense substitutions has low power to detect the effects of selection acting on multiple mutations or standing genetic variation, because in such cases diversity is not necessarily reduced \cite{innan2004,messer2013}. 
Such ``soft sweeps'' are still expected to occur more frequently in functional regions of the genome and could provide an alternative explanation for the observed reduction of diversity in genes. 
To test this possibility, we performed a genome-wide scan for selection using the H12 statistic, a method expected to be sensitive to both hard and soft sweeps \cite{garud2015}. 
Qualitative differences between maize and teosinte remained unchanged even after removing genes in the top 20\% of H12 (Figure \ref{sFig:H12}A).
%\mbh{This makes me wonder about the top 20\%. Is the pattern of diversity reduction different in this fraction of the genome?  If so, that seems like something that should be mentioned.}
%TB Says: Sites in the top 20% of H12 are unlikely to be under background selection, so maize vs teo background selection pattern at these sites seems strange to investigate.
We interpret these combined results as suggesting that purifying selection has left a more pronounced signature in the teosinte genome due to the increased efficacy of selection resulting from differences in effective population size.

\begin{figure*}[tb]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative diversity versus distance to nearest gene in maize and teosinte. 
Shown are \textbf{A} pairwise nucleotide diversity and \textbf{B} singleton diversity.  
Relative diversity is calculated compared to the mean diversity in windows $\geq 0.01 cM$ or $\geq 0.02 cM$ from the nearest
gene for pairwise diversity and singletons, respectively. 
  Lines depict cubic smoothing splines with smoothing parameters chosen via generalized cross validation and shading depicts bootstrap-based 95\% confidence intervals.
  Inset plots depict a smaller range on the x-axis. \label{fig:purify}
  }
\end{figure*}

\subsection{Population expansion leads to stronger purifying selection in modern maize} %JRI
Motivated by the rapid post-domestication expansion of maize evident in our demographic analysis, we reasoned that low-frequency --- and thus younger --- polymorphisms might show patterns distinct from pairwise diversity. 
Singleton diversity around missense and synonymous substitutions (Figure \ref{sFig:singleton}) appears nearly identical to results from pairwise diversity (Figure \ref{fig:hardSweeps}), providing little support for a substantial recent increase in the number or strength of hard sweeps occurring in maize.

In contrast, we observe a significant shift in the effects of purifying selection: singleton polymorphisms are more strongly reduced in and near genes in maize than in teosinte (Figure \ref{fig:purify}B), even after downsampling our maize data to account for differences in sample size. This result is the opposite of the pattern observed for $\pi$, where teosinte demonstrated a stronger reduction of diversity in and around genes than did maize. As was the case when we analyzed $\pi$, this relationship remained after we removed the 20\% of genes with the highest H12 values (Figure \ref{sFig:H12}).
Finally, while direct comparison of pairwise and singleton diversity within taxa are consistent with at least some non-equilibrium dynamics in teosinte, as indicated by our coalescent-based demographic analysis, these too reveal much stronger differences in maize (Figure \ref{sFig:singletonPi}).

@@

\subsection{Demography of domestication} %JRI
Although a number of authors have investigated the demography of maize domestication \cite{eyre1998, tenaillon2004, wright2005}, these  efforts relied on data only from genic regions of the genome and made a number of limiting assumptions about the demographic model.  We show that diversity within genes has been strongly reduced by the effects of linked selection, such that even synonymous polymorphisms in genes are not representative of diversity at unconstrained sites. This implies that genic polymorphism data are unable to tell the complete or accurate demographic history of maize, but the rapid recovery of diversity outside of genes demonstrates that sites far from genes can be reasonably used for demographic inference. Furthermore, by utilizing the full joint SFS, we are able to estimate population growth, gene flow, and the strength of the domestication bottleneck without making assumptions about its duration.

One surprising result from our model is the estimated timing of domestication at $\approx 15,000$ years before present.
While this appears to conflict with archaeological estimates \cite{piperno2009}, we emphasize that this estimate reflects the fact that the genetic split between populations likely preceded anatomical changes that can be identified in the archaeological record. 
We also note that our result may be inflated due to population structure, as our geographically diverse sample of teosinte may include populations diverged from those that gave rise to maize.

@@

\subsection{Demography influences the efficiency of purifying selection}
One of our more striking findings is that the impact of purifying selection on maize and teosinte variability has varied through time. Although we observe a more pronounced decrease in $\pi$ around genes in teosinte than maize (Figure \ref{fig:purify}A), we see a greater decrease in singleton diversity in maize than teosinte (Figure \ref{fig:purify}B). Purifying selection is more efficient in larger populations  \cite{kimura1984}, and our observations using $\pi$  are thus consistent with our demographic model and previous work \cite{eyre1998, tenaillon2004, wright2005, ross2009} which shows a smaller long-term effective population size in maize. Similarly, our observations based on singleton diversity, which reflect more recent patterns of selection, again are in agreement with maize and teosinte demography; from the viewpoint of a recent mutation, the rapid expansion of maize post-domestication means that today maize has a vastly larger effective population size than does teosinte.

\section{Demography and selection interact to shape genetic diversity}
Importantly, our results demonstrate that abstractions such as average long-term $N_e$ cannot capture the dynamic relationship between demography and selection. While it is certainly the case that diversity is lost during a bottleneck, it is equally true that from the viewpoint of a mutation that arises later, the bottleneck is irrelevant. Although the relatively straight-forward demographic history of maize risks oversimplifying this observation, it is applicable to any species that has undergone demographic change. For instance, human populations have experienced recent growth much like maize has, and this growth corresponds to a documented excess of rare genetic variants \cite{keinan2012, tennessen2012, coventry2010}. Our results suggest that in such scenarios, ongoing background selection may lead to a dearth of rare variants in and around genes compared to outside of them. This conflicts with an alternative possibility; that sudden expansion should lead to a comparable increase in rare variants genome-wide.

The abundance of rare variants across the genome, and the efficiency of purifying selection against them, has notable implications for genetic load. One possibility is that the excess of rare variants present in expanding populations implies higher genetic load \cite{keinan2012}. However, more efficient purifying selection in larger populations, which we show operates specifically on these rare variants, should decrease genetic load by purging rare deleterious variants. Several studies have demonstrated that despite different demographic histories, African and European human populations have a similar genetic load \cite{do2015,simons2014}, due in part to new deleterious mutations being increasingly rare as population size increases \cite{simons2014, lohmueller2014}. We propose that this observation could also result from more efficient selection against such mutations. This proposition is in agreement not only with our maize observations, but also with the results of \cite{gazave2013}, who showed via simulations that the increased efficiency of selection in a larger population can partially offset the increase in the total number of segregating deleterious variants.

Across species, our results demonstrate that the genetic impacts of major demographic events such as bottlenecks depend not only on population size reductions, but also on subsequent, potentially rapid and recent, expansion that often comes afterwards. Reduced population sizes do lead to a decreased efficiency of selection \cite[e.g]{renaut2015}, which has long-term impacts on diversity measured via $\pi$. But new mutations, which we measure with singleton diversity, are oblivious to past demography, and therefore selection for or against such sites depends primarily on recent demographic processes.
